A Framework for Studying Animal Behavior Using Deep Learning

### 2024-12-26 docs/papers/log_odds/Paper.tex
wip

Machine learning offers a way in which to at least partially address the bespoke nature of traditional Individual-Based Models (IBMs). By framing individual behaviors as probabilistic decisions among consecutive sets of choices each described by a series of features machine learning can be invoked to largely automate the process of building the decision "engine" behind IBM's. Machine learning excels at identifying patterns in high-dimensional data, enabling models to dynamically incorporate a wide range of environmental and behavioral covariates without requiring custom coding for every new hypothesis. Furthermore, by providing a consistent framing, the same systems and tools can be used across a variety of different problems.

@@

This methodology centers on a two-stage cycle of discovery: (1) testing hypotheses about covariates that are believed to predict behaviors and (2) using the results to generate new hypotheses.

In the first stage, we model behaviors as decisions made from a set of choices, each associated with hypothesized covariates. Then, using machine learning, we build a predictive model to evaluate the likelihood of these selections based on the covariates.

@@

The keystone of this entire process becomes having a clear, quick means of producing machine learning models from our framing of decisions and choices and our covariates, an issue which we turn to next.

@@

This growth poses a significant challenge due to the "curse of dimensionality", where the amount of data required to effectively train models grows exponentially with the dimensionality of the input space \citep{curse}.

@@

We can implement this log-odds model using standard probabilistic deep learning techniques by replicating the log-odds computation across all $M$ choices. The outputs are fed into a softmax layer with $M$ units, where the layer's weights are set to the identity matrix and biases are set to zero. Using categorical cross-entropy as the loss function ensures compatibility with standard probabilistic deep learning while enabling us to train the log-odds weights and significantly reduce the problem's dimensionality.

@@

\subsection*{Alternative Approaches to Reducing Dimensionality}
There are various strategies for reducing the dimensionality of one's data. One straightforward method is to limit the choices available. For instance if making decisions among a set of directions, one could reduce the precision of the angles allowed. While this approach reduces the model's specificity, it significantly simplifies the dimensionality of the problem, which is often a key challenge in probabilistic deep learning.

The methodology proposed, however, achieves dimensionality reduction without sacrificing specificity. That said, there are scenarios where some choices are unlikely to be relevant. In such cases, dimensions corresponding to these irrelevant choices can simply be discarded, further simplifying the model. \newline

Another effective strategy to enhance the data set is leveraging the order invariance of choices in traditional probabilistic problem framing. Specifically, the order in which choices are presented to the model should not matter. For instance, whether a particular choice appears in the first or the thirteenth position should have no impact on the model's operation. This property allows for data augmentation by reordering choices.

In essence, for each training example, $M!$ (factorial of the number of choices) augmented examples can be created. While this exponentially increases the size of the data set as the feature space grows, it provides a mechanism to generate additional training examples from the same data.

The issue with this approach is that as your augmented data size grows to match the needs of the greater dimensionality so too does the time required for training. So while the problem remains theoretically possible, the exponential uptick in time complexity poses a significant practical issue. The proposed approach does not suffer from this issue and remains linear in time with the amount of data provided. 
\section*{Application}

@@

We consider a series of tracks from 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) caught and monitored between 2013 and 2022 \citep{tags1} \citep{tags2}. These tracks were obtained from pop-up satellite archival tags which collect temperature, light level, and depth information at specified (sub day) intervals. This data is then passed through a proprietary algorithm from Wildlife Computers to determine likely longitude and latitude during each day of of monitoring \citep{PSAT}. \newline

@@

Over a test/validation/training split of 40, 71 this resulted in 345,800 contrasts of which 269,800 were used in training and the rest in validation.

@@

Three sets models were trained, one including only distance, one with both distance and movement heading, and one with all four features. Note that while the feature dimensions of these models are 1, 2, and 4 respectively, given the maximum number of choices per decision seen was 33 they dimensionality of a standard probabilistic model would've been 33, 66, and 132 representing a large reduction in the dimensionality of our feature spaces.

@@

Models were trained using AWS Batch using Fargate instances of 2 vcpu's and 4 GB of memory. By taking advantage of batch models could be all trained in parallel allowing for short (15-30 minute) turn around times.

@@

In training the metric of interest is the contrasts normalized log probability (C-NLP) - the average log probability per contrast. For an equivalent metric over all of the decisions we also computed the average log probability per decision for each individual and then computed an average over those across individuals (in order to not favor individuals with many decisions). This is the D-NLP reported in the table below.

@@

\subsection*{Model One: Distance as a Key Feature}
Given the observations that most decisions involve an individual staying put (within its h3 cell) and rarely do fish travel to an h3 cell further than 50km away, it was decided that a more reasonable baseline model would be one that included distance traveled as a feature (model 1).

@@

\subsection*{Model Two: Movement Heading}
Beyond this underlying distance feature the interest is in predicting, when fish move, where they will move. Of particular interest was the individuals in the sitka and yakutat groups as these groups had several individuals exhibiting what appeared to be very directed and extensive movement.

Upon studying these individuals it was noted that changes in mixed layer thickness seemed to incite movement and the individuals tending to stay near higher primary productivity. Therefore a model with net primary productivity and mixed layer thickness was of interest.

However before moving to this model we wanted a baseline model that included movement heading as a feature so that we could determine what aspects of the movement were predicted simply by common headings as opposed to movement specifically related to productivity. Model 2 is this distance plus movement heading model.

@@

This southeasterly pattern is maintained throughout figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} and indeed looking at \ref{fig:ll_increase_movement} there are several groups where the majority of individuals suffer a decrease in log likelihood, in all likelihood due to this common direction not being particularly representative of them.

@@

Looking at the step from model 2 to model 3 in figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} we see now that several of the movements that saw a drop in log likelihood as compared to the model 1 baseline are actually seeing a rise in log-likelihood under this new model indicating that the new features are allowing us to swing the preferred movement heading conditioned on the environment.

@@

Looking at the cases in figure \ref{fig:ll_increase} in which we see significant drops in log likelihood or cases where there is no meaningful change we can see that while there is predictive power in these features chosen thus far - especially for the yakutat, sitka, and nanwalek groups - there are still plenty of individuals that can act as exemplars for searching for new behavioral drivers.

For example in figure \ref{fig:chignik_map} even with the addition of the productivity features in model 3 the westerly movements are still poorly predicted and would warrant investigation.

Likewise the unalaska group is just poorly explained in general beyond the baseline given by model 1 and would represent another case study for further investigation. Perhaps by training a model with the same set of features as model 3 but only over this group we would see a significant improvement in the performance of the model which may lead us to look for demographic or geophysical differences that could explain how two different groups of fish are "using" the same covariates so differently.

Clearly though, the models we have allow us to identify and focus upon individuals that so far resist our attempts to explain them.

@@

This investigation demonstrates the power and flexibility of an iterative modeling approach for understanding behavior. By building baseline models and progressively refining them with additional covariates, we identified both predictive patterns and individuals resistant to explanation, which serve as focal points for further hypothesis generation. The efficiency of this process, enabled by quick training times and parallelization, ensures that refinement is limited only by the investigator's ability to propose and test new hypotheses. Crucially, at every stage, the models not only advance understanding but remain actionable for predictions, underscoring their dual utility as tools for both discovery and application.

>>>>

Machine learning offers a way to at least partially address the bespoke nature of traditional Individual-Based Models (IBMs). By framing individual behaviors as probabilistic decisions among consecutive sets of choices each described by a series of features, machine learning can be invoked to largely automate the process of building the decision "engine" behind IBM's. Machine learning excels at identifying patterns in high-dimensional data, enabling models to dynamically incorporate a wide range of environmental and behavioral covariates without requiring custom coding for every new hypothesis. Furthermore, by providing a consistent framing, the same systems and tools can be used across a variety of different problems.

@@

The methodology centers on a two-stage cycle of discovery: (1) testing hypotheses about covariates that are believed to predict behaviors and (2) using the results to generate new hypotheses.

In the first stage, we model behaviors as decisions made from a set of choices, each associated with hypothesized covariates. Then, using deep learning, we build a predictive model to evaluate the likelihood of these selections based on the covariates.

@@

The keystone of this entire process is having a clear, quick means of producing deep learning models from our framing of decisions, choices, and covariates, an issue which we turn to next.

@@

This growth poses a significant challenge due to the "curse of dimensionality", where the amount of data required to effectively train models can grow exponentially with the dimensionality of the input space \citep{curse}.

@@

We can implement this log-odds model using standard probabilistic deep learning techniques by replicating the "log-odds model" weights across all $M$ choices. The outputs are fed into a softmax layer with $M$ units, where the layer's weights are set to the identity matrix and biases are set to zero. Using categorical cross-entropy as the loss function ensures compatibility with standard probabilistic deep learning while enabling us to train the log-odds weights and significantly reduce the problem's dimensionality.

@@

\subsection*{Alternative Approaches to Reducing Dimensionality}
There are other strategies for reducing the dimensionality of one's data. One straightforward method is to limit the choices available. For instance if making decisions among a set of directions, one could reduce the precision of the angles allowed. While this approach reduces the model's specificity, it significantly simplifies the dimensionality of the problem.

The methodology proposed here, however, achieves dimensionality reduction without sacrificing specificity. That said, there are scenarios where some choices are unlikely to be relevant. In such cases, dimensions corresponding to these irrelevant choices can simply be discarded, further simplifying the model.

Another effective strategy to enhance the data set is leveraging the order invariance of choices in the traditional probabilistic problem framing. Specifically, the order in which choices are presented to the model should not matter. For instance, whether a particular choice appears in the first or the thirteenth position should have no impact on the model's operation. This property allows for data augmentation by reordering choices.

In essence, for each training example, $M!$ (factorial of the number of choices) augmented examples can be created.

The issue with this approach is that as your augmented data size grows to match the needs of the greater dimensionality, so too does the time required for training. So while the problem remains theoretically possible, the potential exponential uptick in time complexity poses a significant practical issue. The proposed approach does not suffer from this issue and remains linear in time with the amount of data provided. 
\section*{Application}

@@

To demonstrate this methodology in practice, we consider a series of tracks from 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) caught and monitored between 2013 and 2022 \citep{tags1} \citep{tags2}. These tracks were obtained from pop-up satellite archival tags which collect temperature, light level, and depth information at specified (sub day) intervals. This data is then passed through a proprietary algorithm from Wildlife Computers to determine likely longitude and latitude during each day of of monitoring \citep{PSAT}. \newline

@@

Over a validation/training split of 40, 71 this resulted in 421,800 contrasts of which 269,800 were used in training and the rest in validation.

@@

Three sets models were trained, one including only distance, one with both distance and movement heading, and one with all four features. Note that while the feature dimensions of these models are 1, 2, and 4 respectively, given the maximum number of choices per decision seen was 33, the dimensionality of a standard probabilistic model would've been 33, 66, and 132 representing a large reduction in the dimensionality of our feature spaces.

@@

Models were trained using AWS Batch using Fargate instances of 2 vcpu's and 4 GB of memory. By taking advantage of AWS Batch,  models could be all trained in parallel allowing for short (15-30 minute) turn around times.

@@

In training the metric of interest is the contrasts' normalized log probability (C-NLP) - the average log probability per contrast. For an equivalent metric over all of the decisions we also computed the average log probability per decision for each individual and then computed an average over those across individuals (in order to not favor individuals with many decisions). This is the D-NLP reported in the table below.

@@

\subsection*{Model One: Distance as a Key Feature}
Given the observations that most decisions involve an individual staying put (within its h3 cell) and rarely do fish travel to an h3 cell further than 50km away, it was decided that a more reasonable baseline model than even odds would be one that included distance traveled as a feature (model 1).

@@

\subsection*{Model Two: Movement Heading}
Beyond this underlying distance feature, the interest is in predicting, when fish move, where they will move. Of particular interest was the individuals in the sitka and yakutat groups as these groups had several individuals exhibiting what appeared to be very directed and extensive movement.

Upon studying these individuals it was noted that changes in mixed layer thickness seemed to incite movement and the individuals tended to stay (anecdotally) near higher primary productivity. Therefore a model with net primary productivity and mixed layer thickness was of interest.

However, before moving to this model we wanted a baseline model that included movement heading as a feature so that we could determine what aspects of the movement were predicted simply by common headings as opposed to movement specifically related to productivity. Model 2 is this distance plus movement heading model.

@@

This southeasterly pattern is maintained throughout figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} and indeed looking at \ref{fig:ll_increase_movement} there are several groups where the majority of individuals suffer a decrease in log likelihood, in all likelihood due to this common direction not being representative of them.

@@

Looking at the step from model 2 to model 3 in figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} we see now that several of the movements that saw a drop in log likelihood as compared to the model 1 baseline are actually seeing a rise in log-likelihood under this new model indicating that the new features are allowing us to swing the preferred movement heading in a way that is conditioned on the environment.

@@

Looking at the cases in figure \ref{fig:ll_increase} in which we see significant drops in log likelihood or cases where there is no meaningful change we can see that while there is predictive power in the features chosen thus far - especially for the yakutat, sitka, and nanwalek groups - there are still plenty of individuals that can act as exemplars for searching for new behavioral drivers.

For example in figure \ref{fig:chignik_map}, even with the addition of the productivity features in model 3 the westerly movements are still poorly predicted and would warrant investigation.

Likewise the unalaska group is poorly explained in general and would represent another case study for further investigation. Perhaps by training a model with the same set of features as model 3 but only over this group we would see a significant improvement in the performance of the model which may lead us to look for demographic or geophysical differences that could explain how two different groups of fish are "using" the same covariates so differently.

All in all though, the models we have allow us to identify and focus upon individuals that so far resist our attempts to explain them.

@@

This investigation demonstrates the power and flexibility of an iterative modeling approach for modeling behavior. By building baseline models and progressively refining them with additional covariates, we identify both predictive patterns and individuals resistant to explanation, which serve as focal points for further hypothesis generation. The efficiency of this process, enabled by quick training times and parallelization, ensures that refinement is limited only by the investigator's ability to propose and test new hypotheses. Crucially, at every stage, the models not only advance understanding but remain actionable for predictions, underscoring their dual utility as tools for both discovery and application.
